Harmony: A Verifiable Systems Programming Language

### 2022-01-14 paper/paper.tex
Pass made

\emph{Harmony} is a Python-like programming language designed to teach
systems specification, implementation, and verification techniques to
students who do not have a background in formal methods.
Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, dereferencing dangling
pointers, trying to multiply two lists with one another,
and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\item Memory leaks.
\end{compactitem}

@@

Harmony programs have three sources of non-determinism:
<{choose}> operations that select a value from a set,
\emph{interleaving} due to multi-threading, and \emph{interrupts}.
\autoref{fig:helloworld} demonstrates the <{choose}> and the <{spawn}>
statements.  Since Harmony programs are model-checked and multiple
executions are explored across a finite set of states, the output of a
Harmony program is a finite automaton that describes all possible
outputs of the program.  Groups of statements can be made atomic
using the <{atomically}> keyword.  Harmony supports
waiting until some condition is reached with its <{when}> statement.

@@

Harmony supports continuations as first-class objects, allowing the
state of a thread can be saved in a variable and later restored.
Continuations enable the implementation of specialized thread
schedulers as well as the implementation of thread forking by
reinstantiating a saved thread state multiple times.

@@

This paper provides an overview of the Harmony language and Charm,
its model checker, and how they may be used by programmers to
develop specify and implement reliable concurrent and/or
distributed systems.

@@

Harmony is heavily based on Python to reduce the learning curve for
users.  Many of Python's powerful constructs and operators are
supported, including comprehensions.  Harmony does
not support classes and instead supports memory allocation and
pointers in a way that is familiar to C programmers.  While this
perhaps seems like a step back in language evolution, many
synchronization algorithms in the literature explicitly consider
memory accesses, and perhaps surprisingly this has been a bonus for
analyzing such algorithms with Harmony.  Moreover, the Harmony model
checker can be used to check for memory leaks or access to dangling
pointers.

For example, \autoref{fig:ms} shows a reproduction from the well-known
Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
Harmony version of the same algorithm.
Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
of a memory location, but inherits the same <{q->field}> notation
from C to access the given field in the dictionary pointed to by
pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
Harmony also complaints, correctly, about a data race when
there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
on an empty queue---the implementer should ensure that accesses to the
\texttt{next} field are sequentially consistent.

The Harmony example in \autoref{fig:ms} shows some other language
features in Harmony that are not in Python.
The <{let x = e: S}> statement introduces a new bound variable <{x}>
with value <{e}> that is only valid within the statement block <{S}>.
Also, Harmony does not have \texttt{return}, \texttt{break}, or
\texttt{continue} statements to force the programmer into a purely
nested control flow.  Each method has a variable <{result}> in which
the return value of the method is stored.

@@

For example, \autoref{fig:synch}(a) shows the specification for locks
from the Harmony <{synch}> module that was used in \autoref{fig:ms},
featuring the Harmony <{when}> statement.
\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
method based on a test-and-set primitive, which in turn is
\emph{specified}.
Imprecisely, methods that use the <{atomically}> keyword are specifications.
Whether something is a specification or an implementation is in the
eye of the beholder.
Important, however, is the notion of modularity that is implied here.
Harmony can model check the implementation of \autoref{fig:ms} using
the specification of a lock in \autoref{fig:synch}(a).
Separately, Harmony can check the correctness of the implementation in
\autoref{fig:synch}(b) against the specification in
\autoref{fig:synch}(a).  This modularity allows larger systems to
be model checked in a way that would not be tractable otherwise.

The Harmony distribution comes with a collection of specifications,
each often with multiple implementations.  The lock specification
comes with the spinlock implementation of \autoref{fig:synch}(a), but
also a ticket lock implementation and one based on continuations and
a blocked queue.  The distribution also includes various concurrent
queue implementations, including the lock-free implementation
of~\cite{MS96}.

@@

\autoref{fig:queue}(a) shows the specification of a concurrent
<{queue}> module.  A queue is represented as a list.
The <{enqueue}> method add a value onto the end of the list,
while the <{dequeue}> method extracts the head, both as atomic actions.

To test a queue, we want to perform a set of concurrent <{enqueue}>
and <{dequeue}> operations concurrently.
\autoref{fig:queue}(b) shows how this can be specified in Harmony.
The test program spawns 4 threads that each select an operation
using the nondetermistic <{choose}> statement.  In case <{enqueue}>
is selected, the thread enqueues its identifier <{self}>.  The program
includes a <{print}> statement at the beginning and end of every queue
operation.  Think of the <{print}> statement as \emph{annotations}
that tell Harmony what the interesting events are with respect to
queue operations.

The test program is similar to what a programmer might have written
as a unit test.  The programmer would run the program a bunch of
times and look at the output in order to see if it looks reasonable.
Doing so is, however, difficult and error prone, as the output tends
to be complex and in the mess of many lines of output it is easy
to overlook problems such as two threads dequeuing the same item
or items being dequeued in the wrong order.  The output of concurrent
print statement can be mangled together.  Not only that, but the
programmer would need to run this test program many times and check
each the output each time.  It even is hard to write a program that
checks the output automatically.

This process is automated in Harmony.  The approach is to
run the model checker twice.  The first time, the test program is
run against the specification of the concurrent queue.  Harmony
outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
time, the same test program is run against the implementation of the
concurrent queue, but with the inclusion of the DFA.  The model
checker ends up verifying the Cartesian product of the implementation
and the DFA and ensures that every behavior of the implementation
program (in terms of its print statements) is also a behavior of
the specification program.

@@

The Harmony <{stop x}> expression stops the current thread and
saves its state in the variable <{x}>.  Another thread can
resume the thread using the statement <{go x r}>, which causes
the original thread to resume, with the <{stop x}> expression
evaluating to the value <{r}>.
\autoref{fig:continuation} shows another implementation of
\autoref{fig:synch}(a) using continuations, where blocked
threads are placed on a FIFO queue and released in that order.
Besides synchronization and scheduling, continuations also enable
the implementation of a primitive such as Posix \texttt{fork}.

@@

Correct interrupt handling is another difficult aspect of
systems programming.  Interrupt handlers typically have various
stringent limitations in what they are allowed to do in order
to ensure the absence of race conditions due to re-entrant code.
The Harmony <{trap f(x)}> statement sets an interrupt handler
<{f(x}}> to be called at an unspecified later time.  You can think
of it as setting an unspecified timer, but it can model any type
of interrupt.  In addition, the Harmony <{setintlevel L}> statement
enables or disables interrupts and returns the old interrupt level.
The model checker tries calling the interrupt handler at every
possible point in the execution of the Harmony program.

@@

Harmony can be used to model a distributed system in which the
only shared variable is a set of messages representing the network.
A message is sent by atomically adding a message to the network:
<{def send(m): atomically network |= {m}}> (`<{|}>' is the union
operator in Python).
To receive a message, a program can use the Harmony <{when}>
statement to wait for a message.  Typically, however, the program
does not want to receive just any message, but a message that satisfies
certain conditions such as: wait for the next message that is addressed
to me and has a sequence number larger than 3.
For this, Harmony supports the <{exists clause}> and allows
<{when}> clauses to be nested:
<{atomically when exists m in network when (m.dst == self) and (m.seq > 3): ...}>
With these constructs, Harmony can be effectively used as a
model checker for distributed algorithms.

@@

Harmony has 8 types of values listed in \autoref{fig:values}.
There are no exceptions: a set can contain dictionaries, a dictionary
can map dictionaries to dictionaries, and so on.  Unlike many
object-oriented programming languages, the programmer never
has to specify what it means for two values to be the same or what the
hash of a value is.

@@

The strings <{"example"}> and <{.example}> are the same---the latter
form is convenient for use in dictionaries.
An \emph{address} represents a path in a hierarchy of dictionaries, with
<{None}> being the empty path.
Examples of \emph{program counters} include method constants and program labels.
A \emph{context} represents the state of a thread.  It includes a program
counter, a map of local variable names to values, and a stack of values.
Harmony makes no difference between lists and tuples and implements
both as dictionaries from integers to values.
So <{[7, 4] = (7, 4) = { 0: 7, 1: 4 }}>.
Note that <{[7, 4][0] = 7}>, as expected.
The empty dictionary is represented as either <{()}> or <{[]}>
(<{{}}> is the empty set).

Except for <{choose}>, any operation on values is deterministic and
defined in the formal specification of the Harmony virtual machine.
For example, like Python, Harmony supports merging of dictionaries
using the union operator, but, unlike Python, Harmony defines
what happens when two dictionaries have the same key but different values.
In that case, Harmony will use the maximum value as defined above.
For dictionary  intersection, Harmony uses the minimum value instead.
Doing so makes dictionary union and intersection deterministic and
commutative.  Moreover, Harmony represents multisets as dictionaries
of values to multiplicities.  Using Harmony's rules on dictionary union
and intersection, union and intersection on multisets are correctly
defined.

@@

\paragraph{Stop/Go.}  Harmony supports continuations as a first-class
construct.  To enable this, \texttt{Stop} saves the current context (with
its program counter incremented) in a specified shared location (usually the
tail of a queue) and removes the context from the context bag and active set.
If the context was atomic (which is usually the case), the active set is
replaced with the domain of the context bag.  Even though its atomic counter
is not zero, the stopped context is not considered atomic at this time, but
will resume being atomic once it is scheduled to run.

@@

\paragraph{Print.}  \texttt{Print} pops a value of the stack and increments
the program counter.  In case a behavior automaton has been specified, the
value is applied to the DFA state and checked for validity.
To reduce state explosion, the HVM does not keep track of the value itself
in the state (like keeping the value in a shared list as part of the state).

@@

\emph{Charm}, the Harmony model checker, is an explicit state model
checker that computes a \emph{Kripke structure} (essentially, a graph)
of states reachable
from the initial state in a breadth-first manner, stopping only if
it encounters an error (such as divide-by-zero or a failing assertion)
or until there are no more new states.

There are three sources of nondetermism in Harmony: \texttt{Choose}
operations, thread interleaving, and interrupts.
When it comes to interleaving, an important observation is that
local operations on the state of a context (updates to its program
counter, local variables, or stack) are not visible to other threads.
An important optimization that significantly
reduces the size of the set of generated states is to treat a
sequence of operations as atomic if only its first operation is an
operation that accesses the shared state (\texttt{Load}, \texttt{Store},
\texttt{Del}, or \texttt{Stop}) or a \texttt{Choose} or \texttt{Print}
operation.  We call such operations \emph{breakable}.
Note that this is essentially a form of Partial Order Reduction~\cite{Val91}.

@@

Given a state and a context, Charm starts with applying the instruction
pointed to by the context's program counter to the state.
Then it continues to apply instructions as long as
the active set is the singleton set (such as when the thread is in
atomic mode) or, if not, as long as the next instruction is not
breakable.  Only the resulting state is added to the Kripke structure,
and the edge indicates the starting context.  During this loop, Charm
does check if the same state recurs---if so it aborts
and reports an infinite loop.

@@

About half of the work of the model checking is trivially parallelizable,
allowing for significant speedups on multi-core machines.

@@

The compiler performs some standard optimizations that do not re-order
expression evaluations, such as constant folding and jump threading.
A more interesting optimization is \emph{dead variable elimination}:
local variables that are no longer in use are automatically removed
from the state using the HVM \texttt{DelVar} instruction.  The reason
for this is to reduce state explosion by removing state that is no
longer in use.

@@

After running the Charm model checker, there are two cases to
consider.  If Charm aborted due to some problematic transition,
then Charm reports the shortest path as a so-called \emph{counter-example}.
If not, the resulting graph is analyzed for various other problems,
including deadlock, busy waiting, and executions
that ended in a state that is not considered final according to
behavior DFA if specified.  We detail below how these problems are
found.

@@

Harmony transforms the Kripke structure into a DAG of its strongly
connected components.  The states in the sink components of this
DAG can be considered the terminal states of the Harmony program.
States in which the context bag is empty represent states in which
all threads have terminated.  In the case the context bag is not
empty, we check if there are any non-\emph{eternal} contexts.  If
so, Harmony reports a \emph{deadlock}.  (Harmony programmers can
specify that certain threads are eternal, such as server threads
that operate in an infinite loop waiting for requests.) If there
is no deadlock and a behavior automaton was specified, Harmony also
checks if the automaton is in a terminal state.  If not, a
\emph{behavior violation} is reported.

@@

A thread is considered \emph{actively busy waiting} if it cannot
escape its strongly connected component without another thread
updating the shared state first.  A common example of this is a
synchronization algorithm that checks for a condition in a loop,
acquiring and releasing the lock each time.  Harmony checks each
strongly connected component for the existence of such situations
and reports them.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=.8\textwidth]{diners.png}
\end{center}
\caption{The HTML output of running the Dining Philosophers program}
\label{fig:dinersHTML}
\end{figure*}

@@

Having such behaviors is not always a problem, but it can be.  For
example, a reader/writer lock implemented as a plain lock meets the
specification of a reader/writer lock, but it does not allow multiple
readers in the critical section.  Unfortunately, checking ${\mathcal
Spec} \subset {\mathcal Impl}$ does not scale well and can only be
done for relatively small Kripke structures.

@@

When Charm finds a problematic state or transition, it outputs
a detailed trace that is called a \emph{counter-example}.
Harmony always outputs the shortest such trace.
However, in order to save memory, Charm does not actually keep
the required information during the construction of the Kripke
structure.  In particular, its edges only contain the context
and the choice if any.  Thus, in order to reconstruct the
counter-example, Charm runs the HVM one more time, this time
re-evaluating the edges along the path and outputting the
detailed information about each transition.

@@

If Charm determines a problem, it outputs the shortest counter-example that
it could find.  However, this counter-example is at the level of the HVM,
not the Harmony program that the programmer wrote.  It can be hard to
dig through a list of HVM instructions.
Harmony includes tools to quickly focus on the root cause of the
problem without having to dig through the HVM instructions.
In particular, in case of a problem, Harmony outputs a compact
but detailed and interactive overview of the problem in HTML format that
can be accessed from any web browser.

@@

The little orange boxes represent individual HVM instructions of the
problematic trace.  You can click on any such box to see the state
at that time, or you can navigate in various ways (forwards one step,
backwards one step, jump to the next line of Harmony code, or jumps
forwards or backwards to a context switch between threads).  The actual
HVM code is in the lower left of the output.

@@

The Harmony language, at least currently, is not usable for production
code.  One reason for this is the <{atomically}> clause.  On a multi-core
processor it would require an instruction to stop all other cores.
Harmony was never designed to be a language for production code.  Because
it is close to languages such as Python and C, it tends to be fairly
easy to hand-translate code developed in Harmony to a desired target
programming language.  However, such hand-translation could easily
introduce a bug that did not exist in the Harmony code.

@@

Promela~\cite{SPIN} is a programming language that serves as input
to the SPIN model checker.  Promela has many of the same features
as Harmony and also includes explicit message channels.  Promela
does not have a call stack and does not support procedure calls,
making it hard to do modular development and verification.

>>>>

\emph{Harmony} is a Python-like programming language designed to teach
systems specification, implementation, and verification techniques to
students who do not have a background in formal methods.
Harmony programs, which can be multi-threaded
or distributed, are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, dereferencing dangling
pointers, trying to multiply two lists with one another,
and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen in the specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: shared variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\item Violations of interrupt safety: a race condition caused by the execution
of an interrupt handler.
\item Memory leaks.
\end{compactitem}

@@

Harmony programs have three sources of non-determinism:
<{choose}> operations that select a value from a set,
\emph{interleaving} due to multi-threading, and \emph{interrupts}.
\autoref{fig:helloworld} demonstrates various non-deterministic
Harmony programs.
Since Harmony programs are model-checked and all possible
executions are explored across a finite set of states, the output of a
Harmony program is a finite automaton that describes all possible
outputs of the program.  Groups of statements can be made atomic
using the <{atomically}> keyword.  Harmony supports
waiting until some condition is reached with its <{when}> statement.

@@

Harmony supports continuations as first-class objects, allowing the
state of a thread to be saved in a variable and later restored.
Continuations enable the implementation of specialized thread
schedulers as well as the implementation of thread forking by
reinstantiating a saved thread state multiple times.
Other uses include exception handlers, coroutines, and generators.

@@

This paper provides an overview of the Harmony language and Charm,
its model checker, and how they may be used by programmers to
specify and implement reliable concurrent and/or distributed systems.

@@

Harmony is heavily based on Python to reduce the learning curve for
users.  Many of Python's powerful constructs and operators are
supported, including comprehensions.  Harmony does
not support classes and instead supports memory allocation and
pointers in a way that is familiar to C programmers.  While this
perhaps seems like a step back in language evolution, many
synchronization algorithms in the literature explicitly consider
memory accesses, and this has been a bonus for
analyzing such algorithms with Harmony.  Moreover, the Harmony model
checker can be used to check for memory leaks or access to dangling
pointers.

For example, \autoref{fig:ms} shows a reproduction of the well-known
Michael\&Scott two-lock concurrent queue~\cite{MS96} along with a
Harmony version of the same algorithm.
Harmony uses the `<{?}>' instead of the `<{&}>' symbol to take the address
of a memory location, but inherits the same <{q->field}> notation
from C to access the given field in the dictionary pointed to by
pointer <{q}>.  <{q->field}> is shorthand for <{(!q)["field"]}> in
Harmony, where `<{!}>' is the Harmony operator to dereference a pointer.
Harmony also complains, correctly, about a data race when
there is a concurrent \texttt{enqueue} and \texttt{dequeue} operation
on an empty queue---the implementer should ensure that accesses to the
\texttt{next} field are sequentially consistent.

The Harmony example in \autoref{fig:ms} shows some other language
features in Harmony that are not in Python.
The <{let x = e: S}> statement introduces a new bound variable <{x}>
with value <{e}> that is only valid within the statement block <{S}>.
Harmony does not have \texttt{return}, \texttt{break}, or
\texttt{continue} statements to force the programmer into a purely
nested control flow.  Each method has a variable <{result}> in which
the return value of the method is stored.

\begin{figure*}
\begin{center}
\begin{tabular}{l|c|c|c|c}
           & <{NOPS=1}> & <{NOPS=2}> & <{NOPS=3}> & <{NOPS=4}> \\ \hline\hline
queue specification    & 11/0.3 & 110/0.3 &  1,286/0.4 &  16,332/0.9 \\ \hline
queue impl. w/o check  & 20/0.5 & 383/0.6 &  8,874/0.8 & 261,878/12.7 \\ \hline
queue impl. with check & 20/0.5 & 425/0.6 & 10,977/1.0 & 437,226/23.6 \\
\end{tabular}
\end{center}
\caption{The number of states generated by the model checker and the
execution time (in seconds) for running \autoref{fig:queue}(b).
This is done for three configurations, varying the value of <{NOPS}>:
the specification (\autoref{fig:queue}(a)), a simple linked list
implementation using locks, and the same implementation combined with
the behavior automaton generated by the first configuration.}
\label{fig:qcmp}
\end{figure*}

@@

For example, \autoref{fig:synch}(a) shows the specification for locks
from the Harmony <{synch}> module that was used in \autoref{fig:ms},
featuring the Harmony <{when}> statement.
\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
method based on a test-and-set primitive, which in turn is
\emph{specified}.
Imprecisely, methods that use the <{atomically}> keyword are specifications.
Whether something is a specification or an implementation is in the
eye of the beholder.
Important, however, is the notion of modularity that is implied here.
Harmony can model check the implementation of \autoref{fig:ms} using
the specification of a lock in \autoref{fig:synch}(a).
Separately, Harmony can check the correctness of the implementation in
\autoref{fig:synch}(b) against the specification in
\autoref{fig:synch}(a).

This modularity allows larger systems to be model checked in a way
that would not be tractable otherwise.
\autoref{fig:qcmp} shows some measurements taken from running
Harmony against the test program in \autoref{fig:queue}(b).
The Harmony compiler allows both constants and modules to be
specified on the command line, so the test program can easily be
run with different versions of the <{queue}> module and different
values for <{NOPS}>, the number of threads spawned to choose and
perform a queue operation.  The table has a row for three different
configurations and a column for four different values for <{NOPS}>.
It is clear that running the test program against the specification
involves significantly fewer states, and therefore less time, than
running the test program against an implementation, particularly
if the behavior of that implementation must be checked as well.

With Harmony, one can check the implementation of a lock separately,
but use the specification of the lock in the implementation of the
queue.  Similarly, one can check the implementation of a queue,
but then use the specification of the queue in a larger application.
Doing so is essentially a form of \emph{Partial Order Reduction}~\cite{Val91},
replacing the <{enqueue}> and <{dequeue}> implementations
by atomic actions.

The Harmony distribution comes with a collection of specifications,
each often with multiple implementations.  The lock specification
comes with the spinlock implementation of \autoref{fig:synch}(a), but
also a ticket lock implementation and one based on continuations and
a queue of blocked threads.  The distribution also includes various concurrent
queue implementations, including the lock-free implementation
of~\cite{MS96}.

@@

\autoref{fig:queue}(a) shows the specification of a concurrent
<{queue}> module.  A queue is represented as a list.
The <{enqueue}> method appends a value onto the end of the list,
while the <{dequeue}> method extracts the head, both as atomic actions.

To test a queue, we want to perform a set of concurrent <{enqueue}>
and <{dequeue}> operations concurrently.
\autoref{fig:queue}(b) shows how this can be specified in Harmony.
The test program spawns 4 threads that each select an operation
using the nondetermistic <{choose}> statement.  In case <{enqueue}>
is selected, the thread enqueues its identifier <{self}>.  The program
includes a <{print}> statement at the beginning and end of every queue
operation.  Think of the <{print}> statements as \emph{annotations}
that tell Harmony what the interesting events are with respect to
queue operations.

The test program is similar to what a programmer might have written
as a unit test.  The programmer would run the program a bunch of
times and look at the output in order to see if it looks reasonable.
Doing so is difficult and error prone, as the output tends
to be complex and in the mess of many lines of output it is easy
to overlook problems such as two threads dequeuing the same item
or items being dequeued in the wrong order.  The output of concurrent
print statement can be mangled together.  Not only that, but the
programmer would need to run this test program many times and check
the output each time.  It even is hard to write a program that
checks the output automatically.

This process is automated in Harmony.  The approach is to
run the model checker twice.  The first time, the test program is
run against the specification of the concurrent queue.  Harmony
outputs a DFA such as shown in \autoref{fig:helloworld}.  The second
time, the same test program is run against the implementation of the
concurrent queue, but with the inclusion of the DFA.  The model
checker ends up verifying the Cartesian product of the implementation
and the DFA and ensures that every behavior of the implementation
program (in terms of its print statements) is also a behavior of
the specification program.  In case the implementation has a behavior
that is not allowed by the specificatino, the model checker can
provide the shortest counter-example.

@@

The Harmony <{stop x}> expression stops the current thread and saves
its state in the variable <{x}>.  Another thread can resume the
thread using the statement <{go x r}>, which causes the original
thread to resume, with the <{stop x}> expression evaluating to the
value <{r}>.  \autoref{fig:continuation} shows another implementation
of the lock specification of \autoref{fig:synch}(a) using continuations,
where blocked threads are placed on a FIFO queue and released in
that order.  Besides synchronization and scheduling, continuations
also enable the implementation of a primitive such as Posix
\texttt{fork}.

@@

Correct interrupt handling is another difficult aspect of
systems programming.  Interrupt handlers typically have various
stringent limitations in what they are allowed to do in order
to ensure the absence of race conditions due to re-entrant code.
The Harmony <{trap f(x)}> statement sets an interrupt handler
<{f(x)}> to be called at an unspecified later time.  You can think
of it as setting an unspecified timer, but it can model any type
of interrupt.  In addition, the Harmony <{setintlevel L}> statement
enables or disables interrupts and returns the old interrupt level.
The model checker tries calling the interrupt handler at every
possible point in the execution of the Harmony program.

@@

Harmony can be used to model a distributed system in which the
only shared variable is a set of messages representing the network.
A message is sent by atomically adding a message to the network:
<{def send(m): atomically network |= {m}}> (`<{|}>' is the union
operator in Python).
To receive a message, a program can use the Harmony <{when}>
statement to wait for a message.  Typically, however, the program
does not want to receive just any message, but a message that satisfies
certain conditions such as: \emph{wait for the next message that is addressed
to me and has a sequence number larger than 3}.
For this, Harmony supports the <{exists}> clause and allows
<{when}> clauses to be nested:
<{atomically when exists m in network when (m.dst == self) and (m.seq > 3): ...}>
With these constructs, Harmony can be effectively used as a
model checker for distributed algorithms.

@@

Harmony has 8 types of values listed in \autoref{fig:values}.
There are no exceptions: a set can contain dictionaries, a dictionary
can map dictionaries to dictionaries, and so on.  Unlike many
object-oriented programming languages including Python, the programmer never
has to specify what it means for two values to be the same or what the
hash of a value is.

@@

The strings <{"example"}> and <{.example}> are the same---the latter
form is convenient for use in dictionaries.
An \emph{address} represents a path in a hierarchy of dictionaries, with
<{None}> being the empty path.
Examples of \emph{program counters} include method constants and program labels.
A \emph{context} represents the state of a thread.  It includes a program
counter, a map of local variable names to values, and a stack of values.
% Harmony makes no difference between lists and tuples and implements
% both as dictionaries from integers to values.
% So <{[7, 4] = (7, 4) = { 0: 7, 1: 4 }}>.
% Note that <{[7, 4][0] = 7}>, as expected.
% The empty dictionary is represented as either <{()}> or <{[]}>
% (<{{}}> is the empty set).

Except for <{choose}>, any operation on values is deterministic and
defined in the formal specification of the Harmony virtual machine.
For example, like Python, Harmony supports merging of dictionaries
using the union operator, but, unlike Python, Harmony defines
what happens when two dictionaries have the same key but different values.
In that case, Harmony will use the maximum value according to the order
defined above.
For dictionary  intersection, Harmony uses the minimum value instead.
Doing so makes dictionary union and intersection deterministic and
commutative.  Moreover, Harmony represents multisets as dictionaries
of values to multiplicities.  Using Harmony's rules on dictionary union
and intersection, union and intersection on multisets are correctly
defined.

@@

\paragraph{Stop/Go.}  Harmony supports continuations as a first-class
construct.  To enable this, \texttt{Stop} saves the current context
(with its program counter incremented) in a specified shared location
(such as the tail of a scheduling queue) and removes the context
from the context bag and active set.  If the context was atomic
(which is usually the case), the active set is replaced with the
domain of the context bag.  Even though its atomic counter is not
zero, the stopped context is not considered atomic at this time,
but will resume being atomic once it is scheduled to run.

@@

\paragraph{Print.}  \texttt{Print} pops a value of the stack and increments
the program counter.  In case a behavior automaton has been specified, the
value is applied to the DFA state and checked for validity.
To reduce state explosion, the HVM does not keep track of the value itself
in the state (like keeping the value in a shared list as part of the state).
Instead, Harmony keeps printed values on the edges between the states
in the Kripke structure.

@@

\emph{Charm}, Harmony's concurrent model checker, is an explicit state model
checker that computes a \emph{Kripke structure} (essentially, a graph)
of states reachable
from the initial state in a breadth-first manner, stopping only if
it encounters an error (such as divide-by-zero or a failing assertion)
or until there are no more new states.

There are three sources of nondetermism in Harmony: \texttt{Choose}
operations, thread interleaving, and interrupts.
When it comes to interleaving, an important observation is that
local operations on the state of a context (updates to its program
counter, local variables, or stack) are not visible to other threads.
An important optimization that significantly
reduces the size of the set of generated states is to treat a
sequence of operations as atomic if only its first operation is an
operation that accesses the shared state (\texttt{Load}, \texttt{Store},
\texttt{Del}, or \texttt{Stop}) or a \texttt{Choose} or \texttt{Print}
operation.  We call such operations \emph{breakable}.
Note that this optimization is essentially Partial Order
Reduction~\cite{Val91}.

@@

Given a state and a context, Charm starts with applying the instruction
pointed to by the context's program counter to the state.
Then it continues to apply instructions as long as
the active set is the singleton set (such as when the thread is in
atomic mode) or, if not, as long as the next instruction is not
breakable.  Only the resulting state is added to the Kripke structure,
and the edge indicates the starting context.  During this loop, Charm
does check if the same state recurs---if so, it aborts
and reports an infinite loop.

@@

In practice, about half of the work of the model checking is trivially
parallelizable, allowing for significant speedups on multi-core machines.

@@

The compiler performs some standard optimizations that do not re-order
expression evaluations, such as constant folding and jump threading.
An unusual optimization is \emph{dead variable elimination}:
local variables that are no longer in use are automatically removed
from the state using the HVM \texttt{DelVar} instruction.  The reason
for this is to reduce state explosion by removing state that is no
longer in use.

@@

After running the Charm model checker, there are two cases to
consider.  If Charm aborted due to some problematic transition,
then Charm reports the shortest path as a so-called \emph{counter-example}.
If not, the resulting graph is analyzed for various other problems,
including deadlock, busy waiting, and executions
that ended in a state that is not considered final according to the
behavior DFA if specified.  We detail below how these problems are
found.

@@

Harmony transforms the Kripke structure into a DAG of its strongly
connected components.  The states in the sink components of this
DAG can be considered the terminal states of the Harmony program.
States in which the context bag is empty represent states in which
all threads have terminated.  In the case the context bag is not
empty, we check if there are any non-\emph{eternal} contexts.
(Harmony programmers can
specify that certain threads are eternal, such as server threads
that operate in an infinite loop waiting for requests.) If
so, Harmony reports a \emph{deadlock}. If there
is no deadlock and a behavior automaton was specified, Harmony also
checks if the automaton is in a terminal state.  If not, a
\emph{behavior violation} is reported.

@@

A thread is considered \emph{actively busy waiting} if it cannot
escape its strongly connected component without another thread
updating the shared state first.  A common example of this is a
synchronization algorithm that checks for a condition in a loop,
acquiring and releasing the lock each time.  Harmony checks each
strongly connected component for the existence of such situations
and reports the shortest example.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=.9\textwidth]{diners.png}
\end{center}
\caption{The HTML output of running the Dining Philosophers program}
\label{fig:dinersHTML}
\end{figure*}

@@

Having such behaviors is not always a problem, but it can be.  For
example, a reader/writer lock implemented as a plain lock meets the
specification of a reader/writer lock, but it does not allow multiple
readers in the critical section.  Unfortunately, checking ${\mathcal
Spec} \subset {\mathcal Impl}$ does not scale well in general and can
only be done for relatively small Kripke structures.

@@

When Charm finds a problematic state or transition, it outputs
a detailed trace that is called a \emph{counter-example}.
Harmony always outputs the shortest such trace.
For ease of debugging, traces are ordered first by the number of
context switches they make, and then by the number of states on
the path.
However, in order to save memory, Charm does not actually keep
the required information during the construction of the Kripke
structure.  In particular, its edges only contain the context
and the choice if any.  Thus, in order to reconstruct the
counter-example, Charm runs the HVM one more time, this time
re-evaluating the edges along the path and outputting the
detailed information about each transition.

@@

If Charm determines a problem, it outputs the shortest counter-example that
it could find.  However, this counter-example is at the level of the HVM,
not the Harmony program that the programmer wrote.  It can be hard to
dig through a list of HVM instructions.
Harmony includes tools to quickly focus on the root cause of the
problem without having to look at the HVM instructions themselves.
In particular, in case of a problem, Harmony outputs a compact
but detailed and interactive overview of the problem in HTML format that
can be accessed from any web browser.

@@

The little orange boxes represent individual HVM instructions of the
problematic trace.  You can click on any such box to see the state
at that time and the HVM instruction it is executing,
or you can navigate in various ways (forwards one step,
backwards one step, jump to the next line of Harmony code, or jumps
forwards or backwards to a context switch between threads).  The actual
HVM code is in the lower left of the output.

@@

The Harmony language, at least currently, is not usable for production
code.  One reason for this is the <{atomically}> clause.  On a multi-core
processor it would require an instruction to stop all other cores.
Harmony was not designed to be a language for production code.  Because
it is close to languages such as Python and C, it tends to be fairly
easy to hand-translate code developed in Harmony to a desired target
programming language.  However, such hand-translation could easily
introduce a bug that did not exist in the Harmony code.

@@

Promela~\cite{SPIN} is a programming language that serves as input
to the SPIN model checker.  Promela has many of the same features
as Harmony and also includes behavior checking.  Promela
does not have a call stack and does not support procedure calls,
making it hard to do modular development and verification.
